Economic Theory of Fertility Explains Modern Natural Selection

### 2020-11-18 why-negative-selection.Rmd
Limit to over-45s. Run with_partner regression by sex.

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are larger among low-income and less educated people, younger parents, and people not living with a partner. Overall effect sizes are substantially bigger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

n_in_regs <- function (var, dv = "n_children") {
  # all PGS have the same NA pattern, so we just use one
  tbl <- table(famhist[
                  ! is.na(famhist[dv]) & ! is.na(famhist$whr_combined), 
                  var
                ])
  c(tbl)
}

@@

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Reweight other regressions (e.g. the omnibus regression in the appendix)?
* Look at geography, esp. in siblings regressions.
* We say we don't see change over time. But, we do see the effect of
  controlling for AFLB changing over generations.
* Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
  his effect sizes compare?
  - he also worries about the quality/quantity tradeoff, which might mean that
    people with fewer children have more grandchildren (than you'd expect).
    As we have the two generations, we can test this. (Actually, there's a
    positive correlation of about 0.1)
* Another point is that negative selection only among poor will increase
    genetic inequality.
* When we have better definition of "partner", bring it across from the other
  project to define with_partner
* If you take "year of last job started" and see if it is after/before
  the first child, you can figure out whether mothers keep working.
  - better still, look at all jobs started and see if any are within e.g. 5 years of
  first child's birth
  - this could then be another way to get at income vs substitution effects. You'd
  predict negative selection effects among mothers who worked. Of course the
  choice to work is endogenous.
* Put mediating effects of pay and education in there (see explore-sex-vs-pay.R)
  - in a multivariate regression, most effects vanish once we control for these, including EA3

@@

## THOUGHTS
* Should we distinguish between men and women in couples?
  - within couples, male h should increase fertility (via earnings)
  - female h should decrease it (ditto?)
  - see explore-sex-vs-pay.R
"
```

@@

As yet we know little about the *mechanisms* behind natural selection in
contemporary human populations. This study uses data from UK Biobank to learn
more. We test for natural selection on 33 different polygenic scores. We extend
the analysis over two generations, using data on respondents' number of siblings
as well as their number of children. This is interesting because consistent
natural selection over multiple generations could lead to substantive effects in
the long run. Most importantly, we examine reproductive rates in different
subgroups of the population, in order to uncover patterns that can help
illuminate the mechanisms behind modern natural selection.

We use data from UK Biobank. We compute 33 polygenic scores, normalized
to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order
of 5% of a standard deviation.

@@

Selection effects are also different between men and women (Appendix
Figure \@ref(fig:plot-sexes)). Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness
and personality traits are more selected for (or less against) among
women, including major depressive disorder (MDD), schizophrenia and
neuroticism, while extraversion is more selected for among men. Waist
circumference and waist-hip ratio are less selected for among women, and
educational attainment is more selected against, though the difference
is only significant for EA3.

@@

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Number of partners", "Household type", "Age at first live birth"), fig.ncol = 1, out.height = c("35%", "27%", "27%"), fig.align = "center"}

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = `N partners`, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        my_theme(legend.position = "bottom") +
        my_labs(x = "", y = "") + 
        scale_y_reordered()

drake::loadd(res_with_partner)
res_with_partner %<>% filter(grepl(":", term))

n_regs <- nrow(res_with_partner)
hhtype_n <- n_in_regs("with_partner")

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner"),
        with_p    = as.character(grepl("TRUE", term)),
        Household = sprintf("%s (N = %s)", Household, 
                      prettyNum(hhtype_n[with_p], big.mark = ","))
      ) %>% 
      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)

@@

[^educ-exception]: Educational attainment is an exception. It is selected 
against in the parents' generation, but selected for if we control for AFLB,
which is the same pattern as for respondents.

@@

This theory predicts that genetic variants will be selected for (or against) in
proportion to their effect on earnings. Figure
\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
that PGS's correlation with earnings in a respondent's first job[^earnings-est],
and against its correlation with educational attainment, a predictor of lifetime
earnings. The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. Thus, the labour market appears to
play an important role in natural selection. Substitution effects
dominate income effects overall, in accordance with the widespread association 
between income and lower fertility [@becker1960economic;@jones2006economic].

@@

We also re-estimate equation \@ref(eq:regression) for each
PGS, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$, suggesting that
earnings and education indeed mediate the effect of PGS on fertility.

@@

Lastly, it is often assumed that the substitution effect will dominate for lone
parents, or those in unstable relationships, since they have less opportunity to
share childcare responsibilities, while the income effect will dominate for
couples who are able to reap the gains from specialization. Indeed, US fertility
decreases faster with education among single mothers than married mothers
[@baudin2015fertility].If so, then earnings-increasing genetic variants may
decrease fertility more among single parents than among couples, as in our data.
The same idea could explain our results for lifetime number of sexual partners,
if this is associated with relationship stability.

@@

Previous work has documented natural selection in modern populations on variants
underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that natural selection effects are highly
concentrated among specific subgroups of the population, including poorer
people, less educated people, younger parents and those with more lifetime
sexual partners. Indeed, among older parents selection effects are actually
reversed. Furthermore, the size of selection effects correlates with the
association between a PGS and labour market earnings. The economic theory of
fertility provides a parsimonious explanation for this.

An alternative theory is that traits selected for are linked to
externalizing behaviour and low time discount rates, via the channel of
early sexual behaviour [@Mills2020.05.06.081273]. The data here provide
some support for this: scores which might plausibly be linked to
externalizing behaviour, like ADHD and age at smoking initiation, are
selected for. However, this theory is less good at explaining variation
in selection across the full range of scores, including physical measures, e.g.
waist-hip ratio and BMI. Externalizing behaviour also does not explain
why selection effects change sign among older parents. Overall, we believe
that the economic theory is the leading explanation.

PGS which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to poor health, or would be seen as undesirable to
have by many people. For example, most people would probably prefer to have high
educational attainment, a low risk of major depressive disorder, and a low risk
of coronary artery disease, but natural selection is pushing against these
traits. Potentially, this could increase the health burden on modern societies,
but that depends on effect sizes. Our results suggest that naïve estimates can
be affected by sample ascertainment bias. This problem may be less serious in
surveys which aim to be representative (as the UK Biobank does not).
Nevertheless, not all respondents consent to the collection of genetic data: for
instance, completion rates for genotype data in the US Health and Retirement
Study were around 80-85% [@hrs2020]. Researchers should be aware of the risks
of ascertainment when studying modern natural selection.

@@

Research on natural selection is controversial. It brings to mind eugenics,
which has been used to justify mass murder and human rights violations, in the
past and possibly the present [@mao1998chinese; @sleeboom2011genetic;
@zenz2020]. Whilst we claim scientific legitimacy for our empirical results, we
have no special authority to speak on how society should react to
them. This is a matter for open debate, in which scientists take part as
equals among many. We therefore choose not to discuss policy implications here.

>>>>

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Overall effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

n_in_regs <- function (var, dv = "n_children") {
  fh_subset <- famhist
  if (dv == "n_children") fh_subset %<>% filter(age_at_recruitment >= 45)
  # all PGS have the same NA pattern, so we just use one
  tbl <- table(fh_subset[
                  ! is.na(fh_subset[dv]) & ! is.na(fh_subset$whr_combined), 
                  var
                ])

  c(tbl)
}

@@

* Check each result with a manually done regression!
* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Reweight other regressions (e.g. the omnibus regression in the appendix)?
* Look at geography, esp. in siblings regressions 
  - maybe Townshend f.189? Not sure essential
* Another point is that negative selection only among poor will increase
    genetic inequality.
* When we have better definition of "partner", bring it across from the other
  project to define with_partner
  - `with_partner` is fine, but if you want to have pairs who had the same children
  you need to do more
* If you take "year of last job started" and see if it is after/before
  the first child, you can figure out whether mothers keep working.
  - better still, look at all jobs started and see if any are within e.g. 5 years of
  first child's birth
  - this could then be another way to get at income vs substitution effects. You'd
  predict negative selection effects among mothers who worked. Of course the
  choice to work is endogenous.
* double check which scores are reverse coded!

@@

## THOUGHTS
* Should we distinguish between men and women in couples?
  - within couples, male h should increase fertility (via earnings)
  - female h should decrease it (ditto?)
  - see explore-sex-vs-pay.R

## RIVAL MECHANISMS
* Lynn 2011 mentions two:
  - social class gradient on ability to delay gratification
  - modern society protects the weak (Darwin 1871, Spencer 1874)
    NB this latter doesn't predict more fertility among "dysgenic" ppl, just
    equal fertility and then genetic drift does its work
    - examples include modern sanitation, post-1800
    - and monogamy!
"
```

@@

As yet we know little about the *mechanisms* behind natural selection in
contemporary human populations. This study uses data from UK Biobank to learn
more. We test for natural selection on 33 different polygenic scores. We extend
the analysis over two generations, using data on respondents' number of siblings
as well as their number of children. This is interesting because consistent
natural selection over multiple generations could lead to substantive effects in
the long run.

Most importantly, we examine reproductive rates in different
subgroups of the population, in order to uncover patterns that can help
illuminate the mechanisms behind modern natural selection. We find that 
across the board, polygenic scores have stronger relationships with fertility 
in specific subgroups of the population. Furthermore, in some subgroups, 
the direction of relationships is actually reversed.

These patterns can be explained by economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Our
results suggest that the substitution effect dominates for single parents,
while the income effect is stronger for couples.

# Data and methods

We use data from UK Biobank. Our sample consists of white British (as defined
by genetic estimated ancestry). For regressions on number of children we 
limit the sample to people over 45, since most fertility is completed by this age.

We compute 33 polygenic scores, normalized
to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order
of 5% of a standard deviation.

@@

Selection effects are also different between men and women (Appendix
Figure \@ref(fig:plot-sexes)). Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness
and personality traits are more selected for (or less against) among
women, including major depressive disorder (MDD), schizophrenia and
neuroticism, while extraversion is more selected for among men. Waist
circumference and waist-hip ratio are less selected for among women, and
educational attainment is more selected against, though the difference
is only significant for EA3. One possible reason for these sex differences is
that polygenic scores may affect fertility via mate search.

@@

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Lifetime number of sexual partners", "Household type", "Age at first live birth, females"), fig.ncol = 1, out.height = c("32%", "32%", "25%"), fig.align = "center"}

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(),
                  legend.box.just      = "left",
                  legend.justification = "left",
                  legend.text          = element_text(size = 7),
                  legend.title         = element_text(size = 8)
                )
sexes_guides <- guides(
                  # fill  = guide_legend(title.position = "top"),
                  # color = guide_legend(title.position = "top")
                )

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = `N partners`, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        sexes_guides

drake::loadd(res_with_partner_sex)
res_with_partner_sex %<>% filter(grepl(":", term))

n_regs <- as.double(nrow(res_with_partner_sex))

res_with_partner_sex %>% 
      mutate(
        score_name = pretty_names(score_name),
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = Household, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        sexes_guides

@@

[^educ-exception]: Educational attainment is an exception. It is selected 
against in the parents' generation, but selected for if we control for AFLB,
which is the same pattern we observe among the respondents' generation.

```{r calc-siblings-children}
sib_chn_cor <- cor(famhist$n_sibs, famhist$n_children, use = "complete")
```

Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
number of children and number of grandchildren (for the parents' generation). We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is `r sib_chn_cor`.

@@

This theory predicts that genetic variants will be selected for (or against) in
proportion to their effect on earnings. Figure
\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
that PGS's correlation with earnings in a respondent's first job[^earnings-est],
and against its correlation with educational attainment, a predictor of lifetime
earnings. The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. Thus, the labour market appears to
play an important role in natural selection. Substitution effects
dominate income effects overall, which fits the empirical association 
between income and lower fertility [@becker1960economic;@jones2006economic].

@@

We also re-estimate equation \@ref(eq:regression) for each
PGS, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
earnings and education indeed mediate the effect of PGS on fertility.

@@

Lastly, it is often assumed that the substitution effect will dominate for lone
parents, or those in unstable relationships, since they have less opportunity to
share childcare responsibilities, while the income effect will dominate for
couples who are able to reap the gains from specialization. Indeed, US fertility
decreases faster with education among single mothers than married mothers
[@baudin2015fertility]. This can explain why earnings-increasing genetic variants 
decrease fertility more among single parents than among couples. The same logic 
could explain our results for lifetime number of sexual partners, if this is 
associated with relationship stability.

@@

Previous work has documented natural selection in modern populations on variants
underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that natural selection effects are highly
concentrated among specific subgroups of the population, including poorer
people, less educated people, younger parents and those with more lifetime
sexual partners. Indeed, among older parents selection effects are actually
reversed. Furthermore, the size of selection effects on a polygenic score
correlates with that score's association with labour market earnings. The 
economic theory of fertility provides a parsimonious explanation for this.

An alternative theory is that traits selected for are linked to
externalizing behaviour and low time discount rates, via the channel of
early sexual behaviour [@Mills2020.05.06.081273]. The data here provide
some support for this: scores which might plausibly be linked to
externalizing behaviour, like ADHD and age at smoking initiation, are
selected for. However, this theory is less good at explaining variation
in selection across the full range of scores, including physical measures, e.g.
waist-hip ratio and BMI (see also appendix Figure \@ref(fig:plot-together)). 
Externalizing behaviour also does not explain
why selection effects change sign among older parents. Overall, we believe
that the economic theory is the leading explanation.

PGS which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to poor health, or are what many people would see
as undesirable to have. For example, most people would probably prefer to have
high educational attainment, a low risk of major depressive disorder, and a low
risk of coronary artery disease, but natural selection is pushing against
genetics linked to these traits. Potentially, this could increase the health
burden on modern populations, but that depends on effect sizes. Our results
suggest that naïve estimates can be affected by sample ascertainment bias. This
problem may be less serious in surveys which aim to be representative (as the UK
Biobank does not). Nevertheless, not all respondents consent to the collection
of genetic data: for instance, completion rates for genotype data in the US
Health and Retirement Study were around 80-85% [@hrs2020]. Researchers should be
aware of the risks of ascertainment when studying modern natural selection.

@@

Any model of fertility is implicitly a model of natural selection. But so far,
the economic and human genetics literatures have developed in parallel.
Integrating the two could deepen our understanding of modern natural selection.
Economics possesses a range of theoretical models on the effects of skills,
education and income [see @hotz1997economics, @lundberg2007american]. One
perennial problem is how to test these theories in a world where education,
labour and marriage markets all interact. Genetic data, such as polygenic
scores, could help to pin down causality. Conversely, theory and empirical
results from economics can shine a light on the mechanisms behind natural
selection.
